Estimating duration in the presence of misclassification

### 2023-12-07 cis-imperfect-testing.tex
First version of full text on imperfect testing

Applying the method of \cref{E-perf-test} to CIS data gives implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
In particular, it estimates a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work~\autocite[e.g.:][]{cevikShedding}.

@@

Based on \cref{imperf-test:fig:bounding-cis-sensitivity}, I propose the following model for test sensitivity, consisting of a linearly declining period then a constant period:
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
\end{equation}
where $t$ is the time in days since infection.
Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
\todo{Figure comparing the variable test sensitivity to fixed and CIS data}

@@

I repeated the simulation study in \cref{E-perf-test:sec:sim-study}, except I used the model of \cref{imperf-test:sec:modelling} with either a constant test sensitivity or the more realistic varying sensitivity.
I focus on the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}.
These were shown to be the best performing priors in \cref{E-perf-test:sec:results}

@@

@@

\subsection{Methods}

\todo[inline]{Check if I've talked about the data before - previous chapter?}

I use the CIS data described in \cref{E-intro:sec:cis}, with positive tests heuristically divided into episodes.
Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \cref{E-episode-def}).

I filter the episodes as follows.
\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
\end{itemize}

I form a prior for $\ntot$ from an extension of the model of \textcite{birrellRealtime}, used to produce incidence estimates.
This gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$.

\subsection{Results}

\todo[inline]{notes on real results into prose}

\begin{itemize}
    \item not sensitive to survival prior
    \item sensitive to test sensitivity assumptions and prior on number missed
    \item comparison with ATACCC
    \item mean and median summaries
    \item 
\end{itemize}

@@

@@

@@

\todo[inline]{write conclusion}

>>>>

Applying the method of \cref{E-perf-test} to CIS data gives implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
In particular, it estimated a much shorter duration of positivity than in \cref{E-ATACCC}, despite those estimates being much more reliable over the first 2--3 weeks.

@@

Based on \cref{imperf-test:fig:bounding-cis-sensitivity}, I propose the following model for test sensitivity, consisting of a linearly declining period then a constant period:
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
  \label{imperf-test:eq:variable-test-sensitivity}
\end{equation}
where $t$ is the time in days since infection.
Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
\todo{Figure comparing the variable test sensitivity to fixed and CIS data}

@@

I repeated the simulation study in \cref{E-perf-test:sec:simulation-study}, except I used the model of \cref{imperf-test:sec:modelling} with either a constant test sensitivity or the more realistic varying sensitivity.
I focus on the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}.
These were shown to be the best performing priors in \cref{E-perf-test:sec:results}

@@

The results when a varying test sensitivity (\cref{imperf-test:eq:variable-test-sensitivity}) is used for the simulation are similar to a constant 0.8 test sensitivity (see \cref{imperf-test:fig:variable-test-sensitivity}).
This suggests that the simplified model, with constant test sensitivity, is sufficient for recovering the true survival time.
Therefore, I will apply this model to the real CIS data in the next section.
\begin{figure}
  % \makebox[\textwidth][c]{
    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-variable-sensitivity}
  \caption[Simulation study results with varying test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a variable test sensitivity.
    All simulations use a constant test sensitivity of 0.8 and the combination prior (as in \cref{imperf-test:fig:misspecified-test-sensitivity}), but the model assumes different values.
    The posterior estimate using a model with a constant test sensitivity of 0.8 (middle panel) is similar to the true value (black line).
  }
  \label{imperf-test:fig:variable-test-sensitivity}
\end{figure}

@@

This section applies the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.

Unlike in the simulation studies, an uninformative prior on $\ntot$ led to implausible estimates of the duration distribution.
The uninformative prior led to high posterior estimates, and hence an implausibly large number of episodes with durations of less than five days.
Therefore, I based an informative prior for $\ntot$ on pre-existing estimates of the total number of infections over the period with posterior mean \numprint{4136368} and standard deviation \numprint{27932}~\autocite{birrellRTM2}.
% This model gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
\todo{Insert correct citation for RTM paper 2 once available}
Approximating this distribution as a negative binomial and scaling the mean to the size f the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$.

With this prior, the model produces plausible estimates of the duration distribution (see \cref{imperf-test:fig:cis-estimates}).
This estimate has more long episodes than the estimate in \cref{E-ATACCC}.
The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for $\lambda$.
However, the survival proportion over the first 4 weeks is sensitive to these choices.
The estimate using a test sensitivity of 0.8 and $\NBc(\numprint{25132}, \numprint{22047})$ give a median survival time most similar to that of \cref{E-ATACCC}.
\todo[inline]{Insert fig showing final estimate and comparison to ATACCC}
\todo[inline]{
  Insert figs showing, for different assumption:
  proportion of individuals still positive at day 30 or 50 (tail),
  overall survival curve, and
  median survival time (in comparison to ATACCC).
}

The assumption that the estimates are most sensitive to is the strength of the prior on $\ntot$.
A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see\todo{figure showing $r$ vs $\ntot$}).
Taking a prior from \textcite{birrellRTM2} means that the posterior survival is close to the \cref{E-ATACCC} estimate in the region where that estimate is reliable.
This justifies the use of this prior.

@@

This chapter shows that a simple model of false negatives can be incorporated into the model of \cref{E-perf-test}.
This model allows estimation of the duration distribution tails without the strong model assumptions used in \cref{E-ATACCC}.
The qualitative features of the estimated distribution are robust to the assumed value for $\psens$, and the choice of prior for $\lambda$, although the quantative details are not.
The estimates of the tail, the primary purpose of this chapter, are in addition robust to the choice of prior for $\ntot$.
However, the rest of the distribution is sensitive to this choice.

@@

An informative prior on $\ntot$ is required is for the bulk of the distribution to be plausible.
In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
One hypothesis is that the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity} causing an unintuitive pattern of results.
In particular, the test sensitivity for long episodes may become much lower than 50\%.
This allows many more episodes to be missed than the model assumes, and hence a higher $\ntot$ is required to explain the data.
Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to when the simulating using a too low test sensitivity.
Violating these assumptions would lead to unpredictable inference results, perhaps those seen here.

Ideally, the test sensitivity would be estimated from the data.
However, this would require a model for time-varying test sensitivity to be incorporated into the model.
Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite something on this}.

The assumption, made in \cref{E-perf-test}, was that the episode start times are independent.
However, because infections are the result of a counting process with intensity shared between individuals (see \cref{E-inc-prev:sec:infection-process}), this assumption is violated.
Prevalence was fairly constant over the period I chose, suggesting the same is true of the incidence.
Additionally, some simulations exploring the results of changing incidence suggested this would not have a large effect on the results (not shown).

% However, comparing the CIS and simulated datasets do not show any differences which suggest what in the data is causing these differences.
% Then, the additional missed infections implied by a 
% This would lead to more missed infections than the model assumes, and hence a higher $\ntot$ is required to explain the data.

The estimates I give in this chapter are the first estimates of the duration of PCR-positivity, including the distribution tails, estimated from general population data.
This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting PCR test results.
Producing these estimates required applying previous survival analysis frameworks in the novel context of the CIS.
The framework needed further modification to account for the presence of false negatives, done in this chapter.

@@

This chapter has produced estimates of the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
The posterior distribution for the survival is concentrated, producing narrow credible intervals.
In the following chapters, I will take the posterior mean for the survival time at each point, neglecting this uncertainty, as the basis for estimating transmission.
First, I will take a backcalculation approach (\cref{E-backcalc}), based on the framework in \cref{E-inc-prev}.
Then, I will take a mechanistic approach (\cref{E-SEIR}), modelling the transmission process explicitly.
